\bf Estimating the number of clusters using cross-validation

### 2016-06-17 cvclust.tex
finish revising section 8

@@

@@

\begin{table}
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table:enrich} Biological process enrichment within gene clusters}
\footnotesize
\begin{tabular}{ccll}
\toprule
Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
\midrule
$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
    &       & ribosomal subunit biogenesis $(73/138)$ & $7.7\times10^{-17}$ \\
    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
$5$ & $517$ & chromosome segregation($53/106$)        & $6.0\times10^{-15}$ \\
    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
\bottomrule
\end{tabular}
\end{center}
\hspace{0.5in} \footnotesize {}
\end{table}

@@

To better compare our results with those of \cite{tavazoie1999systematic}, we
also prepared confusion Table \ref{table:confusion} and Figure
\ref{fig:5_clusters}. The number in cell $(i,j)$ of Table
\ref{table:confusion} shows the number of observations that belong to both the
$i$th cluster of \cite{tavazoie1999systematic}'s and the $j$th cluster of
ours. The plots in Figure \ref{fig:5_clusters} have similar meaning -- the
mini plot on first row and the $4$th column is the mean gene expression level
of observations belong to both the first cluster of
\cite{tavazoie1999systematic}'s and the $4$th cluster of ours. We only plot
the mean expression level if the number of observations in that cell is
greater than $20$. Although \cite{tavazoie1999systematic} found $30$ clusters,
not all of them were discussed in \cite{tavazoie1999systematic}. We only
compare our results with those clusters that were profiled in
\cite{tavazoie1999systematic}. The complete confusion matrix is in Appendix
\ref{app:confusion}.

Cluster~1 mainly consists of observations that did not get profiled in
\cite{tavazoie1999systematic}, as we can see from Table \ref{table:confusion}
and Figure \ref{fig:5_clusters}.

Genes in \cite{tavazoie1999systematic}'s cluster $4$ and $8$ are heavily
concentrated in this our Cluster~2, which also explains the coincidence between the
enriched categories in our cluster $2$ and the functional groups in cluster
$4$ and $8$ of    \cite{tavazoie1999systematic}.

Comparing cluster $3$ to the clusters
found in \cite{tavazoie1999systematic} from Table \ref{table:confusion} and
Figure \ref{fig:5_clusters}, one can see that it mainly consists of genes from
cluster $3$, $7$ and $14$ of \cite{tavazoie1999systematic}. Note that the
cluster $7$ and $14$ are both periodic clusters highlighted in
\cite{tavazoie1999systematic}.

Our Cluster~4 corresponds to Tavazoie's cluster $1$, the most notable functional grouping cluster of
\cite{tavazoie1999systematic}. Indeed, one can see from Table
\ref{table:confusion} that almost entire cluster $1$ of
\cite{tavazoie1999systematic} is contained in our cluster $4$.

Table \ref{table:confusion} shows
that the entire cluster $2$ of \cite{tavazoie1999systematic} is contained in
our cluster $5$, along with some genes from cluster $14$. Note that both
cluster $2$ and $14$ are profiled periodic clusters in
\cite{tavazoie1999systematic}. We can also see from Figure
\ref{fig:5_clusters} that our cluster $5$ groups together genes with similar
cycle phase in \cite{tavazoie1999systematic}'s cluster $2$ and $14$.

In conclusion, our clusters group together genes that participate in the
common biological processes. Our results are broadly align with the results in
\cite{tavazoie1999systematic}, which clustered the data with $k=30$. In
particular, the periodic cell cycle related genes are concentrated in cluster
$5$ and cluster $3$, which consistent with the findings in
\cite{tavazoie1999systematic} (cluster $2$, $7$ and $14$). Indeed, more than
$70\%$ genes which function in mitotic cell cycle process are concentrated in
these two clusters. The number of clusters found by our method is much smaller
than the number of clusters used in \cite{tavazoie1999systematic} and
\cite{dortet2008model}. However, both studies found that most of their
clusters have no characteristics, suggesting overestimate of the underlying
diversity of biological expression classes in the data set
\citep{tavazoie1999systematic}. From Figure \ref{fig:5_clusters}, one can see
clusters in \cite{tavazoie1999systematic} with similar behavior are merged in
our cluster, which leads to smaller number of clusters. Our result is more
parsimonious, while still able to identify the functional grouping of genes in
clusters highlighted in previous studies.

	\begin{figure}[H]
		\centering
	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
	\caption{Yeast data set mean expression profiles. The $5$ clusters found by Gabriel method are on the bottom; Clusters profiled in \cite{tavazoie1999systematic} are on the right.}
	\label{fig:5_clusters}
	\end{figure}

\begin{table}[H]
\caption{\label{table:confusion} \small{Confusion matrix comparing $5$ clusters to clusters profiled in \cite{tavazoie1999systematic}}}
\centering
\begin{tabular}{lcccccr}
  \hline
 & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
  \hline
Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
   \hline
\end{tabular}
\end{table}

>>>>

\begin{figure}
		\centering
	\includegraphics[width=5.7in, height=5.7in]{wei/split_plots.pdf}
	\caption{Yeast data set mean expression profiles. The $5$ clusters found by
  the Gabriel method are on the bottom; Clusters profiled in
  \cite{tavazoie1999systematic} are on the right.}
	\label{fig:5_clusters}
\end{figure}

@@

\begin{table}
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table:enrich} Biological process enrichment within gene clusters}
\footnotesize
\begin{tabular}{ccll}
\toprule
Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
\midrule
$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
    &       & ribosomal subunit biogenesis ($73/138$) & $7.7\times10^{-17}$ \\
    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
$5$ & $517$ & chromosome segregation ($53/106$)       & $6.0\times10^{-15}$ \\
    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
\bottomrule
\end{tabular}
\end{center}
\hspace{0.5in} \footnotesize {}
\end{table}

@@

@@

In the \citet{tavazoie1999systematic} analysis, those authors performed
$k$-means clustering with $k = 30$; they found 23 of the clusters to be
uninterpretable, and they found 7 clusters to be meaningful. To compare our
clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
comparing our clusters with the 7 interpretable Tavazoie clusters in
Table~\ref{table:confusion}.  (Appendix~\ref{app:confusion} contains a
confusion matrix for all 30 Tavazoie clusters, including the non-interpretable
ones.) Entry $(i,j)$ of the confusion matrix gives the number of genes in
Tavazoie's Cluster~$i$ and our Cluster~$j$.

\begin{table}
\caption{\label{table:confusion} Confusion matrix comparing
    the 5 clusters found by Gabriel cross-validation to the 7 interpretable
    clusters found by the \cite{tavazoie1999systematic} analysis}
\centering
\begin{tabular}{lcccccr}
  \toprule
 & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
  \midrule
Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
  \midrule
  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
   \bottomrule
\end{tabular}
\end{table}

Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
$(i,j)$ of the upper left part of this figure gives the mean expression level
for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
the plots in the margins give the mean expression levels for Tavazoie's
clusters (top right) and our clusters (bottom left). In
Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
number of genes in that cell is greater than $20$.

Our Cluster~1 mainly consists of genes that Tavazoie et al.\ found to be in
uninterpretable clusters. Our Cluster~2 contains high concentrations of
Tavazoie's Clusters~4 and~8. Our first periodic cluster, Cluster~3, contains
high concentrations of Tavazoie's Clulsters~3,~7, and~14; this is notable,
because Tavazoie et al.\ highlighted their Clusters~7 and~14 as being
periodic. Our Cluster~4 contains almost all of Tavazoie's Cluster~1, along
with part of Tavazoie's Cluster~4. Finally, our second periodic cluster,
Cluster~5, contains almost all of Tavazoie's Cluster~2, along with part of
Tavazoie's Cluster 14; this, again, is notable, because Tavazoie et al.\
highlited these clusters as being periodic.

For the clusters that Tavazoie et al.\ were able to characterize, our analysis
broadly agrees with the earlier clustering. The major difference between our
analysis and that of \citet{tavazoie1999systematic} is that we are able to
identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
instead of $k = 30$), and we are able to interpret all of the clusters
found by our analysis.
